---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# polygonoverlap

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/benmarwick/polygonoverlap.svg?branch=master)](https://travis-ci.org/benmarwick/polygonoverlap)
[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/polygonoverlap/master?urlpath=rstudio)
<!-- badges: end -->

The goal of polygonoverlap is to compute the probability that an observed area of overlap between two sets of polygons is due to chance. We assume one set of polygons can freely move, and another set is fixed in space. We randomly relocate the first set of polygons many times, and for each relocation, we compute the area of overlap. After repeating this random relocation many times, we generate a distribution of overlap areas. We compare this distribution of randomly-generated overlap areas to the observed overlap area to compute a p-value. This is useful for answering the question "is the location of the first set of polygons non-random with respect to overlapping the fixed set of polygons?", or "is the observed amount of overlap of the two sets of polygons due to chance alone, or some other process?"

## Installation

You can install the development version of polygonoverlap from GitHub with:

``` r
remotes::install_github("benmarwick/polygonoverlap")
```

## Example

Let's load the library and some polygon shapefiles to work with. Currently this works with shapefiles that are loaded into R using `rgdal::readOGR()` to produce a `SpatialPolygonsDataFrame` object. 

```{r load-libs, message=FALSE}
library(polygonoverlap)
library(here)
```

```{r load-data, message=FALSE}
# load the data contained in the pkg so we can demonstrate
bounding_box_polygon <- rgdal::readOGR(here("data-raw/MK_II_excv_outline.shp"), verbose = FALSE)
input_polygons <- rgdal::readOGR(here("data-raw/rock_poly.shp"), verbose = FALSE)
other_polygons <- rgdal::readOGR(here("data-raw/skele_poly.shp"), verbose = FALSE)
```

We can take a quick look to see these polygons, green is our `input_polygons` which we assume can be freely moved, and red is our `other_polygons`, which we assume are fixed:

```{r}
sp::plot(bounding_box_polygon)
sp::plot(input_polygons, add = TRUE, border = "green")
sp::plot(other_polygons, add = TRUE, border = "red")
title(main = "Plot of input polygons")
```

Notice that the red polygons overlap themselves a bit. In our function below we dissolve those overlap areas into single polygons so we don't measure multiple overlaps in one location. 

Now we shuffle the set of `input_polygons` to random locations within our bounding box area. We repeat this random shuffle many times, and save the locations of the polygons for each random shuffle event:

```{r cache=TRUE}
# This may take a minute or two
n <- 1000
input_polygons_randomly_shuffled <- 
  shift_poly_to_random_points(bounding_box_polygon,
                              input_polygons,
                              n)
```

For each random shuffle event, we compute the size of the area of intersection (or overlap) between our `input_polygons` and another set of polygons (we do not shuffle this other set):

```{r cache=TRUE}
areas_of_overlap_from_random_shuffle <- 
  compute_overlap_area_of_polygons_randomly_shuffled(input_polygons_randomly_shuffled, 
                                                     other_polygons)
```

We can make a plot of this distribution:

```{r}
library(ggplot2)
ggplot(areas_of_overlap_from_random_shuffle,
       aes(area)) +
  geom_histogram() + 
    labs(x = expression("Areas of intersection of our two sets of polygons (m"^2*")"),
         y = "Frequency") +
  ggtitle(paste0("Distribution of areas of polygon overlap produced by ", n, " random shuffles"))
```

Now we can compare this randomly generated distribution to the observed overlap of our two sets of shapefiles:

```{r}
observed_polygon_overlap <- 
  compute_overlap_area_of_polygons_observed(input_polygons, 
                                            other_polygons)
```

The value we get here is `r round(observed_polygon_overlap,3)` square meteres. We can combine this with our random distribution data to compute a p-value:

```{r}
pval <- 1 - sum(areas_of_overlap_from_random_shuffle$area <= observed_polygon_overlap) / n
```

Using this p-vaue we can say that `r pval*100`% of our randomly-shuffled input shapefiles result in an overlap with the other shapefiles that is equal to or greater than our observed overlap area. This indicates that our observed area of overlap is probably not random but a result of deliberate placement of the input polygons over or near the other polygons. The answer to the question is 'yes, the association of the two polygon sets is non-random'.

And we can show the observed value and p-value on the histogram like this:

```{r}
ggplot(areas_of_overlap_from_random_shuffle,
       aes(area)) +
  geom_histogram() + 
    labs(x = expression("Areas of intersection of our two sets of polygons (m"^2*")"),
         y = "Frequency") +
  ggtitle(paste0("Distribution of areas of polygon overlap produced by ", 
                 n, 
                 " random shuffles")) + 
  geom_vline(xintercept = observed_polygon_overlap, 
             col = "red") + 
  ylim(0, 100) +
  annotate("text", 
           x = 0.45, y = 85, 
           label = paste0("Observed \nvalue = ",
                         round(observed_polygon_overlap,2), 
                         " \n(p = ", round(pval,3), ")"), col = "red")
```

## Background

This method was developed for this publication:

> Lowe, Kelsey M., Lynley A. Wallis, Colin Pardoe, Ben Marwick, Chris Clarkson, Tiina Manne, Mike A. Smith and Richard Fullagar 2014. Ground-penetrating radar and burial practices in western Arnhem Land, Australia. _Archaeology in Oceania_ <https://doi.org/10.1002/arco.5039>

This package is an excerpt of code and data prepared for that paper, which was originally uploaded to  https://github.com/benmarwick/Rocks-and-burials-at-Madjebebe and archived at http://dx.doi.org/10.5281/zenodo.10616

